Testing traversability of protein tunnels with flexible ligands using motion planning

### 2017-06-06 paper.tex
algorithms

\author{Vojt\v ech Von\' asek\inst{1} \and Barbora Kozl\'\i kov\'a\inst{2} \and Adam Jur\v{c}\'\i k\inst{2} \and Martin Saska\inst{1}}
\institute{
Faculty of Electrical Engineering,  
Czech Technical University in Prague, 
Technick\'a 2, 166 27, Prague 6, Czech Republic
\email{vonasek@labe.felk.cvut.cz}
\and
Faculty of Informatics,  
Masaryk University, 
Botanick\'a 68a, 602 00 Brno,
}

@@

\abstract{
%\begin{abstract}
Proteins are involved in many biochemical processes and they function is achieved by interactions with other molecules such as ligands.
The interaction takes place in an active site inside the protein, which is accesible from the surface by so called tunnels.
The knowledge about tunnels can help researches for example to better select target ligand in order to achieve a desired
behavior of the protein.
Widely used tools for tunnel detection are based on Voronoi diagrams.
This however requires to approximate the ligand by a sphere, which is not always realistic.
In this paper, we propose a motion planning technique for analysis of traversability of such tunnels by flexible ligands.
To decrease the dimension of the related configuration space, the flexibility of ligands is modeled by a set of expected conformations.
The proposed method is based on the Rapidly Exploring Random Tree motion planning.
The spheres representing the tunnel are used to guide the growth of the RRT tree in the configuration space.
Something about visualization ...
%\end{abstract}
}

\section{Introduction}

\red{prepsat, todle uz jsme meli vickrat}
Protein structures are essential components of all living organisms.
Proper understanding of their structure and function is important in many fields, including drug design, agriculture, cosmetics, etc.
Although this knowledge is still very hard to reveal, several protein structures have been already investigated in sufficient detail. 
Such investigation can incorporate computational methods which help to analyze the geometry of protein structure,  its shape, surface area, and inner void space. 
These parameters then suggest and influence protein behavior, namely its reactivity with other molecules.

Proteins interact with other molecules (small ligands, ribonucleic acid chains, other proteins, etc.) which can lead to changes in properties of the protein or the interacting molecule.
The chemical reaction between the protein and ligand takes places in so called active site, that is accessible 
from the protein surface by a tunnel.

The knowledge about the tunnels is important in many fields .. todo.
For example, one task in protein engineering is to change selected properties of a protein, e.g. its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
ligand from the outside environment to the active site or vice versa.

The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or 
a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
These methods compute the tunnels assuming a spherical ligand (probe).
The resulting tunnels are represented as a sequence of spheres.
The active sites are usually connected with the protein surface with multiple pathways.
However, not all paths are biochemically relevant, so only paths with a given minimal bottleneck have to be reported (the bottleneck is
the radius of the smallest sphere that can traverse the whole tunnel) 
Bio-chemicists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and its bottleneck.

\red{TODO}
something about screeing: what is it, what's is the input and output, in which applications (research/industry) it is useful.

Such decision is however not precise and it requires previous expertise with the domain.
Of the reasons is that typical ligands are non-spherical and therefore it is difficult to estimate their traversability in tunnels
computed for a spherical probe.
For better decision, it is necessary to consider also the shape of the ligand being investigaed, and consequently, to also
consider rotations of the whole ligand.
Moreover, ligands are rather flexible molecules, so the relative positions of their atoms can change as well.

@@

%hese approaches are not suitable for tunnel detection in dynamic proteins, a
%, which limits the usage of these methods for spherical ligands.
%To increase precision by 10, the number of cells increases by $10^3$.
%Another disadvantage is that the 3D grid does not consider orientation of the ligand and therefore, the computations 
%are useful only for spherical ligands.

@@

Finding tunnels for a general, non-spherical and possibly a flexible ligand can be solved by searching
a feasible path in the corresponding configuration space $\C$.
The dimension of the space is given by the degrees of freedom (6D for a solid ligand and 6D+$n$ for a flexible ligand with
$n$ rotational bonds).
The feasible configurations, where the ligand (robot) does not touch and obstacle, form the free region $\CF$.
The tunnel traversability can be then formulated as a path planning from the start configuration $\qinit \in \CF$ 
to a goal configuration $\qgoal\in\CF$, where both $\qinit$ and $\qgoal$ are located at the begining and the end of the
tunnel, respectively.
Sampling-based methods like PRM~\cite{kavrakiPRM} or RRT~\cite{lavalleRRT} can be used to search such a path.
%Sampling-based techniques like PRM~\cite{kavrakiPRM} or RRT~\cite{lavalleRRT} can be used to search such a high-dimensional space.
%The tunnel detection in dynamic proteins can also be formulated as path planning~\cite{vonasek2016application}. in 4D (i.e., 3D $\times$ time) 
%configuration space assuming a spherical probe.
%Tunnel detection in dynamic proteins leads to path planning 
%in a high-dimensional configuration space (4D for the probe and 7D for the ligand).
%This space can be searched using sampling-based motion planning methods.
%Instead of a systematic search, which would be time and memory consuming, randomized sampling can be used
%to obtain necessary information about the configuration space. %~\cite{Lav06}. 
%Path planning approaches studied in robotics can be used to solve tunnel detection~\cite{guieysse2008structure} as well.
%Sampling-based path planning methods like Probabilistic Roadmaps (PRM)~\cite{kavrakiPRM} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT} are able to find feasible paths for complex objects with arbitrary geometries.
The idea of sampling-based motion planning is to randomly sample the configuration space.
The random samples are classified as free or non-free using collision detection and  the free ones are stored into a roadmap.
A path in the roadmap then represents a motion in the workspace.
%The usage of collision detection for classification of samples allows the application of sampling-based methods for path planning for robots (objects) of arbitrary shapes.
%As the sampling-based methods work in the configuration space, they can cope with objects (robots) of many degrees of freedom
% and arbitrary shapes.
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
% to study proteins, e.g. in  
%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.

Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query motion planning method.
RRT incrementally builds a configuration tree $\T$ rooted at the initial configuration.
In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node in the tree $\qnear \in \T$ is found.
A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$.
If $\qnew$ is collision-free, it is added to the tree.
The algorithm terminates if the tree approaches the goal configuration close enough.

Beyond robotics, sampling-based motion planners have been successfully applied in computational biology in studies
related to 
protein folding~\cite{al2012motion,gipson2012computational,cortes2010simulating,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},analysis of loop motions~\cite{cortes2004geometric}, and even for tunnel detection~\cite{vonasek2016application,vonasek2017tunnel}.
%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%In our previous works, we proposed to employ RRT also for the purp
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.

The most relevant paper is~\cite{guieysse2008structure}, where RRT is used to compute pathways for flexible ligands towards the active site, but
without the possibility to select a preferable path/tunnel.
%This method however does not compute explicitly all tunnels, rather a single pathway.
%This allows us to build a sequence of trees, where each tree corresponds to each frame of protein dynamics.
%The nodes of the trees are connected to their parents in the same frame, or to their parents in the previous frame.
%This allows us to find paths in this sequence of tree between two given frames, that correspond to the dynamic tunnels.
%In comparison to~\cite{guieysse2008structure}, the approach~\cite{vonasek2016application} assumes spherical probes, which is motivated
%by practical reasons.
%Flexible ligands have many degrees of freedom, which increases dimension of the configuration space to be searched and makes
%the collision-detection more complex.
%Moreover, it is required to define suitable metric and possibly also an efficient data structure for nearest-neighbor search.
%Despite the flexibility of ligands, their movements in proteins are strongly limited, as they are typically larger than spherical probes.
%Traversability of the ligand through a protein may depend on its initial orientation, 
% which can further limit the growth of the configuration tree.
%Contrary, configuration space of  spherical probes is only 3D with the possibility to employ Euclidean metric and KD-trees for the
%nearest neighbor search.
%It is more practical to first detect tunnels for a spherical probe and then analyze traversability of the actual shape of ligand.
%In this case, the traversability analysis is limited only to a tunnel and its vicinity, which decreases volume of the configuration space and
%decreases the number of atoms involved in collision detection.

In this paper, we propose to employ motion planning to compute the trajectory for a given ligand in a given tunnel.
This brings several advantages.
From motion planning point of view, the tunnels have properties of narrow passages.
By focusing the motion planning only to the surrounding of a tunnel, the total volume of the configuration space decreases which increases
the relative volume of the free space, which increases probability of placing random sampling in the perspective regions of the configruation 
space.

@@

@@

By reducing the scale of the ligand's atomic radii, the configuration tree can explore more areas of the configuration space.
Such techniques has been used for motion planning of deformable objects, 
e.g.,~\cite{frank2008efficient,bayazit2001ligand,alterovitz2008motion,lamiraux2001flexible,kavraki1998towards,gayle2005path} 
and for motion planning of rigin objects in amongst 
flexible obstacles~\cite{rodriguez2006planning,frank2008efficient,phillips2014representation}.

@@

@@

Ligands are small molecules interacting with the protein. 
Due to the interaction, the relative positions of the ligand's atoms can change.
The ligands can be modeled as a multi-link kinematic chain where the torional angles phi/psi can change in a given range~\cite{songPFpath}.
After the torsional angles are set, 3D position of each atom of the ligand is computed.
By changing the angles, various conformations can be generated.
However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.

Another approach is to model the ligand flexibility using a predefined set of feasible conformations $\L$
To decrease the computational burgen, a set of feasible conformations $\L$ is prepared, e.g. using Rosseta tool.
%The conformations can be prepared using Molecular Dynamics software such as Rosetta.
%For each conformation $c\in \L$ an energy potential is also calculated.
%Let $\L_i$ denote the set of conformations that 
We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
%The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
%can be switched to the confomration $c_j \in C$.

The configuration space is searched using the RRT principle~\cite{lavalleRRT}.
The main loop of the method is described in Alg.~\ref{alg::main}.
In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found
and the node is expanded.
The random samples are not generatd only from whole $\C$, as in the original RRT, but the sampling
is guided using moving virtual goal~\cite{vonasek2009rrt}.
The virtual goal $v$ is the index of the actual sphere of the tunnel, around which the samples
are generated with probability $\gb$. 
After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
The algorithm termines after a predefined planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = N$.

    $S = $expandConfigurations($\qnear,\L$)\;

    \For{$q \in S$}{
        $\T$.addNode($q$)\;
        $\T$.addEdge($\qnear, q$)\;
    }

\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::expand}expand}
\KwIn{
   configuration $\qnear$ to be expanded,
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
\KwOut{
    set $S$ of collision-free configurations reachable from $q$
}
\hrule
\ForEach{$l \in \L$}{
    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin)$}{
        \For{$i = 1,\ldots,m$}{
            $q' = q$\;
            translate $q$ by random around $q'$\;
            rotate randomly \;
            $q'.l = l$\;
            $q'.s = s$\;
            \If{$isCollisionFree(q')$}{
                $S = S \cup \{ q \}$\;
            }
        }
    }
}
\end{algorithm}

The result of the proposed planning method is a tree of collision-free configurations, from which a path leading from $\qinit$
to the nearest node towards $\qgoal$ can be found.
Due to stochastic nature of RRT, it is not guaranteed that the tree always reached the goal configuration $\qgoal$.
Despite, a trajectory leading as close as possible towards $\qgoal$ is extracted fromm the tree.

>>>>

\author{
    Vojt\v ech Von\' asek \and 
    Barbora Kozl\'\i kov\'a  \and 
    Adam Jur\v{c}\'\i k  \and 
    David Bend\'{a}\v{r}  \and
    Martin Saska }

\institute{
    V. Von\'{a}sek, Martin Saska \at 
    Faculty of Electrical Engineering,  
Czech Technical University in Prague, 
Technick\'a 2, 166 27, Prague 6, Czech Republic
\email{vonasek@labe.felk.cvut.cz}  
\and
B. Kozl\'{\i}kov\'{a}, A. Jur\v{c}\'{\i}k \at
Faculty of Informatics,  
Masaryk University, 
Botanick\'a 68a, 602 00 Brno, \\
\and 
\inst{3}
TODO DAVID \at TODO DAVID
}

@@

\abstract{
\red{needs rewrite:
Function and behavior of proteins are achieved by interactions with other molecules such as ligands.
The interaction takes place in an active site inside the protein, which is accessible from the surface by so called tunnels.
The knowledge about tunnels can help researches for example to better select target ligand in order to achieve a desired
behavior of the protein.
Widely used tools for tunnel detection are based on Voronoi diagrams and they compute the tunnels for a virtual spherical probe.
The tunnels are then scored based on their bottleneck, length and also curvature. 
It is however not easy to determine if a given non-spherical ligand can pass such a tunnel.
To provide the chemists better information about the tunnels, the full shape of ligands should be considered.
}
Sampling-based motion planning techniques can be used to compute trajectories for the non-spherical ligands in the protein tunnels.
In this paper, we propose a modification of Rapidly Exploring Random Tree (RRT) to find the trajectories.
To enable the ligands to move in the tunnels, the planning is realized with a reduced-scale of the ligands. 
The flexibility of ligands is achieved by using a predefined set of ligand conformations.
For the purpose of further analysis, we computed trajectories are visualized ..
}

\section{Introduction \& motivation}

\red{this is fast copy and paste from previous paper, please rewrite, cut to at ca. 1 page}

%Protein structures are essential components of all living organisms.
Proper understanding of structure and function of proteins is important in many fields, including drug design, agriculture, cosmetics, etc.
%Although this knowledge is still very hard to reveal, several protein structures have been already investigated in sufficient detail. 
Computational methods are often used to analyze the geometry of protein structure,  its shape, surface area, and inner void space. 
Proteins interact with other molecules (small ligands, ribonucleic acid chains, other proteins, etc.) which can lead to changes in properties of the protein or the interacting molecule.
The chemical reaction between the protein and ligand takes places in so called active site, that is accessible 
from the protein surface by a tunnel.

The knowledge about the tunnels is important in many fields..
For example, one task in protein engineering is to change selected properties of a protein, e.g. its stability under different outer conditions~\cite{Koudelakova2013} or activity of the protein towards the other molecules~\cite{Pavlova2009}.
This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
ligand from the outside environment to the active site or vice versa.

The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or 
a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
These methods compute the tunnels assuming a spherical ligand (probe).
The resulting tunnels are represented as a sequence of spheres.
The active sites are usually connected with the protein surface with multiple pathways.
However, not all paths are biochemically relevant, so only paths with a given minimal bottleneck have to be reported (the bottleneck is
the radius of the smallest sphere that can traverse the whole tunnel) 
Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and its bottleneck.

\red{TODO}
something about screening: what is it, what's is the input and output, in which applications (research/industry) it is useful.

Such decision is however not precise and it requires previous expertise with the domain.
Of the reasons is that typical ligands are non-spherical and therefore it is difficult to estimate their traversability in tunnels
computed for a spherical probe.
For better decision, it is necessary to consider also the shape of the ligand being investigated, and consequently, to also
consider rotations of the whole ligand.
Moreover, ligands are rather flexible molecules, so the relative positions of their atoms can change as well.

@@

@@

Finding tunnels for a general, non-spherical and possibly a flexible ligand can be solved by searching
a feasible path in the corresponding configuration space $\C$.
The dimension of the space is given by the degrees of freedom (6D for a solid ligand and 6D+$n$ for a flexible ligand with
$n$ rotational bonds).
The feasible configurations, where the ligand (robot) does not touch and obstacle, form the free region $\CF$.
The tunnel traversability can be then formulated as a path planning from the start configuration $\qinit \in \CF$ 
to a goal configuration $\qgoal\in\CF$, where both $\qinit$ and $\qgoal$ are located at the beginning and the end of the
tunnel, respectively.
Sampling-based methods like PRM~\cite{kavrakiPRM} or RRT~\cite{lavalleRRT} can be used to search such a path.
The idea of sampling-based motion planning is to randomly sample the configuration space.
The random samples are classified as free or non-free using collision detection and  the free ones are stored into a roadmap.
A path in the roadmap then represents a motion in the workspace.
Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query motion planning method.
RRT incrementally builds a configuration tree $\T$ rooted at the initial configuration.
In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node in the tree $\qnear \in \T$ is found.
A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$.
If $\qnew$ is collision-free, it is added to the tree.
The algorithm terminates if the tree approaches the goal configuration close enough.

Beyond robotics, sampling-based motion planners have been successfully applied in computational biology in studies
related to 
protein folding~\cite{al2012motion,gipson2012computational,cortes2010simulating,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},analysis of loop motions~\cite{cortes2004geometric}, and even for tunnel detection~\cite{vonasek2016application,vonasek2017tunnel}.
%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%In our previous works, we proposed to employ RRT also for the purp
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
% to study proteins, e.g. in  
%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.

The most relevant paper is~\cite{cortes2005path}, where RRT is used to compute pathways for flexible ligands towards the active site, but
without the possibility to select a preferable path/tunnel.

In this paper, we propose to employ motion planning to compute the trajectory for a given ligand in a given tunnel.
This brings several advantages.
From motion planning point of view, the tunnels have properties of narrow passages.
By focusing the motion planning only to the surrounding of a tunnel, the total volume of the configuration space decreases which increases
the relative volume of the free space, which increases probability of placing random sampling in the perspective regions of the configuration 
space.

@@

% in cortes2005path:
%In this paper two kinds of large-amplitude motion are treated: protein loop conformational changes (involving pro-
%tein backbone flexibility) and ligand trajectories to deep active sites in proteins (involving ligand and protein side-chain flex-
%ibility). First studies performed using our two-stage approach (geometric search followed by energy refinements) show that,
%    compared to classical molecular modeling methods, quite   similar results can be obtained with a performance gain of
% several orders of magnitude. Furthermore, our results also  indicate that the geometric stage can provide highly valuable information to biologists.

@@

By reducing the scale of the ligand's atomic radii, the configuration tree can explore more areas of the configuration space.
Such techniques has been used for motion planning of deformable objects, 
e.g.,~\cite{frank2008efficient,bayazit2001ligand,alterovitz2008motion,lamiraux2001flexible,kavraki1998towards,gayle2005path} 
and for motion planning of rigid objects in among 
flexible obstacles~\cite{rodriguez2006planning,frank2008efficient,phillips2014representation}.

@@

Ligands are small molecules interacting with the protein. 
Due to the interaction, the relative positions of the ligand's atoms can change.
The ligands can be modeled as a multi-link kinematic chain where the torsional angles phi/psi can change in a given range~\cite{songPFpath}.
After the torsional angles are set, 3D position of each atom of the ligand is computed.
By changing the angles, various conformations can be generated.
However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.

Another approach is to model the ligand flexibility using a predefined set of feasible conformations $\L$
To decrease the computational burden, a set of feasible conformations $\L$ is prepared, e.g. using Rosseta tool.
%The conformations can be prepared using Molecular Dynamics software such as Rosetta.
%For each conformation $c\in \L$ an energy potential is also calculated.
%Let $\L_i$ denote the set of conformations that 
We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
%The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
%can be switched to the confomration $c_j \in C$.

The configuration space is searched using the RRT principle~\cite{lavalleRRT}.
The main loop of the method is described in Alg.~\ref{alg::main}.
In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found
and the node is expanded.
The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
is guided using moving virtual goal~\cite{vonasek2009rrt}.
The virtual goal $v$ is the index of the actual sphere of the tunnel, around which the samples
are generated with probability $\gb$. 
After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
The algorithm terminates after a predefined planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = N$.

    expand($\qnear$)\;

   % \For{$q \in S$}{
    %    $\T$.addNode($q$)\;
    %    $\T$.addEdge($\qnear, q$)\;
    %}

\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::expand}expand}
\KwIn{
   configuration $\qnear$ to be expanded,
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
%\KwOut{
%    set $S$ of collision-free configurations reachable from $\qnear$\;
%}
\hrule
\ForEach{$l \in \L$}{
%    $newnodes = $ false\;
    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin)$}{
        \For{$i = 1,\ldots,m$}{
            $q' = q$\;
            translate $q$ by random around $q'$\;
            rotate randomly \;
            $q'.l = l$\;
            $q'.s = s$\;
            $qn$ = nearest node in $\T$ towards $q'$\;
            \If{$isCollisionFree(q')$ {\bf and} $d(q',\qnear) < d(q',qn)$}{
                $\T$.addNode($q$)\;
                $\T$.addEdge($\qnear,q$)\;
%                $S = S \cup \{ q \}$\;
%                $newnodes = $ true\;
                {\bf break}\;
            }
        }
    }
}
%\return $S$\;
\end{algorithm}

The result of the proposed planning method is a tree of collision-free configurations, from which a path leading from $\qinit$
to the nearest node towards $\qgoal$ can be found.
Due to stochastic nature of RRT, it is not guaranteed that the tree always reached the goal configuration $\qgoal$.
Despite, a trajectory leading as close as possible towards $\qgoal$ is extracted from the tree.

% ======================================================================
%
%    UNUSED TEXT      UNUSED TEXT      UNUSED TEXT     UNUSED TEXT
% ======================================================================

@@
